---
title: "Lab 3a - Exercises"
author: "Environmental Data Analytics | John Fay and Luana Lima "
date: "Fall 2025"
output: pdf_document
editor_options: 
  chunk_output_type: console
---

## Set up the session
- Load packaged: tidyverse, lubridate, & here
- Import the `EPAair_PM25_NC2019_raw.csv` file
  - Set strings to be read in as factors
  - Set the "Site.ID" column to be read in as a factor
- Set the date column to be a date

```{r Setup}
#Import packages
library(tidyverse); library(lubridate); library(here)

#Check workspace
here()

#Import data
pm25 <- read.csv(
  here('Data/Raw/EPAair_PM25_NC2019_raw.csv'),  #----ENTER THE FILENAME HERE, USING THE here() PACKAGE----
  stringsAsFactors = TRUE,
  colClasses = c("Site.ID"='factor')
)

#Fix the date
pm25$Date <- mdy(pm25$Date) #changed date from factor to date 
class(pm25$Date) #date

```


## Explore properties of the dataframe
-   How many different sites are included in the data set?
-   What is the mean daily PM2.5 concentration?
-   In what units of are PM 2.5 concentrations reported?

```{r Explore the data}
# Compute the number of different sites
unique(pm25$Site.ID) #25 levels

# Compute the mean and median PM 2.5 concentration
mean(pm25$Daily.Mean.PM2.5.Concentration) #7.68
median(pm25$Daily.Mean.PM2.5.Concentration) #7.4
summary(pm25$Daily.Mean.PM2.5.Concentration)

# What are the units
summary(pm25$UNITS) #ug/m3 LC 

# What is the date range (min and max dates)
min(pm25$Date) #2019-01-01
max(pm25$Date) #2019-12-31
summary(pm25$Date)
```


## Plot distribution of values

-   Number of samples using each different source 
-   Distribution of daily mean concentration values
-   Distribution of daily mean concentration, by source*
  *   There are multiple ways to visualize this...

```{r Visualize the data}
#Number of samples using each different source 
ggplot(pm25, aes(x=Source)) + geom_bar()

#Distribution of daily mean concentration values
ggplot(pm25, aes(x=Daily.Mean.PM2.5.Concentration)) + geom_histogram()

ggplot(pm25) + 
  geom_histogram(aes(x = Daily.Mean.PM2.5.Concentration), bins=20)


#Distribution of daily mean concentration, by source

##histogram
ggplot(pm25) +
geom_histogram(aes(x=Daily.Mean.PM2.5.Concentration, fill=Source))

##frequency polygon
ggplot(pm25) + 
  geom_freqpoly(aes(Daily.Mean.PM2.5.Concentration, color = Source))

##frequency polygon
ggplot(pm25) + 
  geom_boxplot(aes(Daily.Mean.PM2.5.Concentration, x=Source))


```

