---
title: "Assignment 2: Coding Basics"
author: "Morgan Finch"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons/labs in Environmental Data Analytics on coding basics.

## Directions

1.  Rename this file `<FirstLast>_A02_CodingBasics.Rmd` (replacing `<FirstLast>` with your first and last name).
2.  Change "Student Name" on line 3 (above) with your name.
3.  Work through the steps, **creating code and output** that fulfill each instruction.
4.  Be sure to **answer the questions** in this assignment document.
5.  When you have completed the assignment, **Knit** the text and code into a single PDF file.
6.  After Knitting, submit the completed exercise (PDF file) to Canvas.

## Basics, Part 1

1.  Generate a sequence of numbers from one to 55, increasing by fives. Assign this sequence a name.

2.  Compute the mean and median of this sequence.

3.  Ask R to determine whether the mean is greater than the median.

4.  Insert comments in your code to describe what you are doing.

```{r basics 1}
#1. 
seq5 <- seq(1, 55, 5) # from 1 to 55 by increments of 5
seq5

#2. 
mean_seq5 <- mean(seq5) #mean is 26
med_seq5 <- median(seq5) #median is 26

#3. 
greater <- mean_seq5 > med_seq5 
greater #False, mean is not greater than the median

```

## Basics, Part 2

5.  Create three vectors, each with four components, consisting of (a) student names, (b) test scores, and (c) whether they are on scholarship or not (TRUE or FALSE).

6.  Label each vector with a comment on what type of vector it is.

7.  Combine each of the vectors into a data frame. Assign the data frame an informative name.

8.  Label the columns of your data frame with informative titles.

```{r basics 2}
vec_names <- c("Morgan", "Clare", "Sophi", "Andrea") #create vector for student names
vec_names #return vec
class(vec_names) #class: character

vec_scores <- c(95, 92, 98, 77) #create vector for test scores
vec_scores
class(vec_scores) #class: numeric

vec_scholar <- c(FALSE, FALSE, TRUE, FALSE) #create vector for scholarship or not
vec_scholar
class(vec_scholar) #class: logical

##create a dataframe
dataframe1 <- data.frame(vec_names,vec_scores,vec_scholar) 
is.data.frame(dataframe1) #confirm dataframe

#Set column names
names(dataframe1) <- c("Student Names","Test Scores","Scholarship Status")

```

9.  QUESTION: How is this data frame different from a matrix?

> Answer: Data frames and matrices are both 2 dimensional structures. Data frames, however, are more general than a matrix and can have different element. A matrix has elements of the same type.

10. Create a function with one input. In this function, use `if`...`else` to evaluate the value of the input: if it is greater than 50, print the word "Pass"; otherwise print the word "Fail".

11. Create a second function that does the exact same thing as the previous one but uses `ifelse()` instead if `if`...`else`.

12. Run both functions using the value 52.5 as the input

13. Run both functions using the **vector** of student test scores you created as the input. (Only one will work properly...)

```{r basics 3}
#10. Create a function using if...else
ifelse_1 <- function(x) {
  if(x > 50) {   #if greater than 50
    x="Pass"     #print pass
  } 
  else {
    x="Fail"     #else print fail
  }
}
less50 <- ifelse_1(20); less50 #20 is less than 50, Fail
greater50 <- ifelse_1(60); greater50 #60 is greater than 50; Pass

#11. Create a function using ifelse()
ifelse_2 <- function(x){
  ifelse(x>50, "Pass", "Fail")
}

value20 <- ifelse_2(20); value20 #fail <50
value75 <- ifelse_2(75); value75 #pass >50
  
#12a. Run the first function with the value 52.5
print(ifelse_1(52.5)) #Pass

#12b. Run the second function with the value 52.5
print(ifelse_2((52.5))) #Pass

#13a. Run the first function with the vector of test scores
###value_1 <- ifelse_1(vec_scores);value #Error in if (x > 50) { : the condition has length > 1

#13b. Run the second function with the vector of test scores
value_2 <- ifelse_2(vec_scores); value_2 #"Pass" "Pass" "Pass" "Pass"

```

14. QUESTION: Which option of `if`...`else` vs. `ifelse` worked? Why? (Hint: search the web for "R vectorization")

> Answer: The if...else function returned the error if (x \> 50) { : the condition has length \> 1. The if...else function can only run for one element at a time. The ifelse() function was able to compute each component in the vector.

**NOTE** Before knitting, you'll need to comment out the call to the function in Q13 that does not work. (A document can't knit if the code it contains causes an error!)
